---
title: "Hepatitis A Report"
author: "Fereidoun Mianji"
date: "10/29/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Introduction:

The effectiveness of public vaccnation on the infection rate of Hepatitis A is studied in the Us by state. This shows the trend of the disease over about half a century. The analysis is based on the weekly statistics collected across the states. As the dataset includes many disease, first we need to filter it to only keep Hepatitis data for the years that a minimum of 30 weeks are repoted for it.

```{r loading libs and saving the disease data, message=FALSE}
library(tidyverse)
library(dslabs)

data(us_contagious_diseases)
the_case <- "Hepatitis A"

HepatitisA_dat <- us_contagious_diseases %>%
  filter(weeks_reporting >= 30) %>%   
  filter(disease == the_case) %>%  
  mutate(disease_rate = count / population * 10^4 * 52 / weeks_reporting) %>%  ## calc. rate per ten tousands 
  mutate(state = reorder(state, disease_rate))
save(HepatitisA_dat, file = 'rdas/HepatitisA_dat.rda')
```

2. Comparing geographical zones:

To see if the behaviour of the disease has been the same in various parts of the country, the data for 6 states are compared. The plot shows that, 2005 can be considered as the breakdown of Hepatitis A in all regions.

```{r Analyzes the data for this disease and compares the trend in the states, message=FALSE}

library(RColorBrewer)


load("rdas/HepatitisA_dat.rda")
the_case <- "Hepatitis A"
HepatitisA_dat %>% filter(state %in% c("Texas", "Florida", "California", "New York",
                      "Washington", "North Dakota") &!is.na(disease_rate)) %>%
  ggplot(aes(year, disease_rate, col = state)) +
  geom_line() +
  ylab("Hepatitis A cases per 10,000 population") + 
  geom_vline(xintercept=2005, col = "blue") +
  ggsave('fig/sample-states-trend.png')
```

3. Comparing all the states in a single plot:
 
Its very informative to compare all the states on a single graph to see if they all have followed the same trend.

```{r to compare all the states with others}
HepatitisA_dat %>% ggplot(aes(year, state, fill = disease_rate)) +
  geom_tile(color = "grey50") +
  scale_x_continuous(expand=c(0,0)) +
  scale_fill_gradientn(colors = brewer.pal(9, "Reds"), trans = "sqrt") +
  geom_vline(xintercept=2000, col = "blue") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        legend.position="bottom",
        text = element_text(size = 8)) +
  ggtitle(the_case, ' cases per 10,000 population') +
  ylab("") + xlab("") +
  ggsave("fig/all-states-trend.png")
```

4. The US average:

And finally, it is shown that the trend of the disease in the US, after an increase between 1965 to 1970, has been steadily declining getting to a minimum in 2010.


```{r to plot us average on state trends}
avg <- us_contagious_diseases %>%
  filter(disease==the_case) %>% group_by(year) %>% 
  summarize(us_rate = sum(count, na.rm = TRUE) /
              sum(population, na.rm = TRUE) * 10000)

HepatitisA_dat %>%
  filter(!is.na(disease_rate)) %>%
  ggplot() +
  geom_line(aes(year, disease_rate, group = state), color = "grey50",
            show.legend = FALSE, alpha = 0.2, size = 1) +
  geom_line(mapping = aes(year, us_rate), data = avg, size = 1) +
  scale_y_continuous(trans = "sqrt", breaks = c(5, 25, 125, 300)) +
  ggtitle("Cases per 10,000 by state and the US average") +
  xlab("") + ylab("") +
  geom_text(data = data.frame(x = 1977, y = 2),
            mapping = aes(x, y, label="US average"),
            color="black") +
  geom_vline(xintercept=2009, col = "blue") +
  ggsave("fig/us-average.png")
```


